```{r echo=FALSE, warning=FALSE, message=FALSE}
#!/usr/bin/Rscript

options(stringsAsFactors=FALSE)
library(knitr)
library(ggplot2)
library(gdata)
library(rgl)
knit_hooks$set(webgl = hook_webgl)

## for debugging
#args=commandArgs(TRUE)
#inFile=args[1]
#sampleFile=args[2]

#inFile <- "./PCA/data_clean.eigenvec"
#sampleFile <- "/home/julius/soft/ref1000G/ALL_1000G_phase1integrated_v3.sample"
```

Report on PCA of merged reference and sample data
========================================================

Report produced by J. Juodakis on **`r format(Sys.time(), "%F %R")`**

PCA eigenvectors obtained from file **`r inFile`**

Sample data obtained from file **`r sampleFile`**

------------------------------------------

```{r read_data, echo=FALSE}
ev=read.table(inFile,h=F)
sample=read.table(sampleFile,h=T)

refs=ev[grepl("ref",ev$V2),]
ours=ev[!grepl("ref",ev$V2),]

ones=rep("sample",dim(ours)[1])
ours2=cbind(ours$V2,ours[,3:5],ones,ones)
names(ours2)=c("iid","p1","p2","p3","population","group")
refs2=cbind(refs$V2,refs[,3:5],sample$population,sample$group)
names(refs2)=c("iid","p1","p2","p3","population","group")

all=rbind(ours2,refs2)
```

```{r calculate_eucl, echo=FALSE}
sube2=all[all$group=="EUR","p1"]
sube3=all[all$group=="EUR","p2"]
sube4=all[all$group=="EUR","p3"]

sube2a=sube2-mean(sube2)
sube3a=sube3-mean(sube3)
sube4a=sube4-mean(sube4)

sde=sqrt(sd(sube2a)^2+sd(sube3a)^2+sd(sube4a)^2)

eucl=sqrt(sube2a^2+sube3a^2+sube4a^2)
eucl2=c(-eucl,eucl)
sdeucl=sd(eucl2)

euclmusu=sqrt((ours2$p1-mean(sube2))^2+(ours2$p2-mean(sube3))^2+(ours2$p3-mean(sube4))^2)
oursE=cbind(ours2,euclmusu)
oursBad=oursE[oursE$euclmusu>3*sdeucl,]
```

## Significant outliers

For outlier detection, Euclidean distance of each sample from the EUR group center of mass is calculated. Distribution of this distance is presented in further below.  
An SD value was calculated after mirroring this distribution around x=0, and 3*SD is used as a cutoff. Individuals whose distances exceed this cutoff are listed in the table below.

```{r EuclData, echo=FALSE}
kable(oursBad[,c(1:4,7)],format="markdown")
hist(euclmusu,col="grey",breaks=30, main="Euclidean distance distribution",xlab="Euclidean distance from EUR group center")
```

------------
## Visualization of the results

The plots below present 2D projections of the points along the first 3 principal components.  
Colors reflect different groups or populations of the reference genome.

```{r flat_plots, echo=FALSE}
tinkami=all[(all$group=="EUR"),]
musu=all[(all$group=="sample"),]

ggplot(all, aes(x=p1, y=p2, color=group))+geom_point(size=3)+ggtitle("p1 vs p2")
ggplot(all, aes(x=p1, y=p3, color=group))+geom_point(size=3)+ggtitle("p1 vs p3")
ggplot(all, aes(x=p2, y=p3, color=group))+geom_point(size=2)+ggtitle("p2 vs p3")

ggplot(all, aes(x=p1, y=p2, color=population))+geom_point(size=2)+ggtitle("p1 vs p2 on POPULATIONS")
```

The plots below show the projections of the points along the first 2 principal components. *Ellipses* show the region between 5-95 percentiles of the data, assuming t or normal distributions. *Circles* show the region of 3 SD of Euclidean distance (assuming mirrored distribution) or 3 times geometrical average of SD along each principal component.  
SD of Euclidean distance: **`r sdeucl`**  
Geometrical average of SD along each PC: **`r sde`**  

```{r ellipse_plots, echo=FALSE}
ggplot(all, aes(x=p1, y=p2, color=group))+geom_point(size=2)+stat_ellipse(data=tinkami)+
        ggtitle("p1 v p2 + t distribution 5-95 %")
ggplot(all, aes(x=p1, y=p2, color=group))+geom_point(size=2)+stat_ellipse(data=tinkami, type="norm")+
        ggtitle("p1 v p2 + normal distribution 5-95 %")

ggplot(all, aes(x=p1, y=p2, color=group,label=iid))+geom_point(size=2)+
        stat_ellipse(data=all[all$group=="EUR",], type="euclid",level=3*sdeucl)+
        ggtitle("p1 v p2 + 3 SDs of euclidean distance")
ggplot(all, aes(x=p1, y=p2, color=group,label=iid))+geom_point(size=2)+
        stat_ellipse(data=all[all$group=="EUR",], type="euclid",level=3*sde)+
        ggtitle("p1 v p2 + 3x geom. averaged SD")
```

The following images represent the positiong of data in 3D space formed by the first 3 principal components. *Sphere* covers the region of 3 SD of Euclidean distance (same as above). *Lines* are the actual Euclidean distance to the EUR group center of mass.  
To analyze the object in 3D, it is best to manually run the .Rmd file in R.

```{r calculate_things, echo=FALSE, warning=FALSE}
all$group=as.factor(all$group)
attach(ours2)
xai=interleave(p1,rep(mean(p1),length(p1)))
yai=interleave(p2,rep(mean(p2),length(p2)))
zai=interleave(p3,rep(mean(p3),length(p3)))
detach("ours2")
```

```{r plot_3d1, webgl=TRUE, echo=FALSE, warning=FALSE}
plot3d(all[,c("p1","p2","p3")],col=as.integer(all$group))
par3d(scale=c(1,1,1), windowRect=c(200,200,800,800))
view3d(theta=0,phi=-45)

grid3d(c("x","y","z"))
spheres3d(mean(sube2),y=mean(sube3),z=mean(sube4),radius=3*sdeucl,front="line",back="line",alpha=0.5)
segments3d(xai,yai,zai)
```

```{r plot_3d2, webgl=TRUE, echo=FALSE, warning=FALSE}
plot3d(all[,c("p1","p2","p3")],col=as.integer(all$group))
par3d(scale=c(1,1,1), windowRect=c(200,200,800,800))
view3d(theta=-60,phi=0)

grid3d(c("x","y","z"))
spheres3d(mean(sube2),y=mean(sube3),z=mean(sube4),radius=3*sdeucl,front="line",back="line",alpha=0.5)
segments3d(xai,yai,zai)
```

```{r plot_3d3, webgl=TRUE, echo=FALSE, warning=FALSE}
plot3d(all[,c("p1","p2","p3")],col=as.integer(all$group))
par3d(scale=c(1,1,1), windowRect=c(200,200,800,800))
view3d(theta=30,phi=0)

grid3d(c("x","y","z"))
spheres3d(mean(sube2),y=mean(sube3),z=mean(sube4),radius=3*sdeucl,front="line",back="line",alpha=0.5)
segments3d(xai,yai,zai)
```

```{r plot_3d4, webgl=TRUE, echo=FALSE, warning=FALSE}
plot3d(all[,c("p1","p2","p3")],col=as.integer(all$group))
par3d(scale=c(1,1,1), windowRect=c(200,200,800,800))
view3d(theta=90,phi=60)

grid3d(c("x","y","z"))
spheres3d(mean(sube2),y=mean(sube3),z=mean(sube4),radius=3*sdeucl,front="line",back="line",alpha=0.5)
segments3d(xai,yai,zai)
```

```{r plot_labels, echo=FALSE, eval=FALSE}
## this clutters the plot images, so use only in R
text3d(oursBad[,2:4],text=oursBad$iid)

## this spins
play3d(spin3d(rpm=3))
```
