```{r echo=FALSE}
#!/usr/bin/Rscript
library(knitr)
options(stringsAsFactors=F)

cutoffs<-read.table(cutoffFile,h=F)
imiss_cutoff= as.numeric(cutoffs[1,])
hribos= as.numeric(cutoffs[2,])
me_cutoff= as.numeric(cutoffs[3,])

## for debugging
#inFile = "./QC_ind_1/starting_stats"
#outDir = "./QC_ind_1/"
#cutoffFile = "./QC_ind_1/shiny_ind_cut.txt"
#imiss_cutoff= 0.02
#hribos= 2
#me_cutoff =100

roundNumber=as.numeric(substr(outDir,nchar(outDir)-1,nchar(outDir)-1))
```

Analysis report of QC individual statistics, round `r roundNumber`
========================================================

Report produced by J. Juodakis on **`r format(Sys.time(), "%F %R")`**

Data statistics obtained from files `r inFile`.imiss etc.

Output written to directory `r outDir`

------------------------------------------

```{r read_stats, echo=FALSE}
## individuals missing
imis=read.table(paste(inFile,".imiss",sep=""),h=T,stringsAsFactors=F)
totalI=dim(imis)[1]

##heterozygosity
het=read.table(paste(inFile,".het",sep=""),h=T,stringsAsFactors=F)
h=(het$N.NM. - het$O.HOM.)/het$N.NM.
hMean=mean(h)
hSD=sd(h)

## mendel errors
merrors=read.table(paste(inFile,".imendel",sep=""),h=T,stringsAsFactors=F)

## initialize tables for output stat files
statsI=imis[,c(2,6)]
statsI=cbind(statsI, h)
statsI=cbind(imis$FID,statsI)
statsI=merge(statsI,merrors, by="IID", sort=F, all=T)
statsI=statsI[,c(1:4,6)]
names(statsI)<-c("FID","IID","F_MISS","HETEROZ", "MERRORS")
```

## General Info

In this round of QC, statistics on **`r totalI` individuals** were analyzed.

```{r do_tests, echo=FALSE}
## tests for merrors and heterozygosity
metest=!(statsI$MERRORS>me_cutoff)
metest=(metest==T | is.na(metest))
hettest=!(statsI$HETEROZ<hMean-hribos*hSD | statsI$HETEROZ>hMean+hribos*hSD)

## updating the statsI
statsI=cbind(statsI, statsI$F_MISS<imiss_cutoff)
statsI=cbind(statsI, hettest)
statsI=cbind(statsI, metest)
names(statsI)<-c("FID","IID","F_MISS","HETEROZ","MERRORS","pass_imiss","pass_het","pass_mendel")
```

The following plots present the filtering criteria and results for 3 filters - proportion of SNPs missing in each individual (IMISS), proportion of heterozygous loci in each individual (HET) and SNPs with Mendelian errors per individual (MERR).

```{r plots1, echo=FALSE, fig.height=12, fig.width=12}
## draw histograms
par(mfrow=c(2,2))
hist1=hist(imis$F_MISS,breaks=100,col="grey",main="IMISS filter",ylab="individuals",xlab="proportion of SNPs missing")
yc1=max(hist1$counts)/2
xc1=mean(range(imis$F_MISS,na.rm=T))
n_excl1=totalI-sum(hist1$counts[1:sum(hist1$breaks<=imiss_cutoff)],na.rm=T)
text(xc1,yc1,paste("selected for exclusion: ",n_excl1,sep="") )  
text(xc1,yc1*1.2,paste("imiss cutoff:",imiss_cutoff))
abline(v=imiss_cutoff, col="red")

hist5=hist(h,breaks=100,col="grey",main="HET filter",ylab="individuals",xlab="proportion of heterozygous loci")
nexcl5=totalI-sum(hettest)
yc5=max(hist5$counts)/2
xc5=mean(range(h,na.rm=T))
text(xc5,yc5,paste("selected for exclusion:",nexcl5))
text(xc5,yc5*1.2,paste("Heterozygosity cutoff in SDs:",hribos))
abline(v=hMean, col="red")
abline(v=hMean+hribos*hSD, col="red")
abline(v=hMean-hribos*hSD, col="red")

hist3=hist(merrors$N,breaks=100,col="grey", main="MERR filter",ylab="individuals",xlab="SNPs with errors in this individual")
yc3=max(hist3$counts)/2
xc3=mean(range(merrors$N,na.rm=T))
n_excl3=sum(hist3$counts)-sum(hist3$counts[1:sum(hist3$breaks<=me_cutoff)],na.rm=T)
text(xc3,yc3,paste("selected for exclusion: ",n_excl3,sep="") )  
text(xc3,yc3*1.2,paste("mendel cutoff:",me_cutoff))
abline(v=me_cutoff, col="red")
```

```{r echo=FALSE}
### final filter
ffI=(statsI$pass_het==TRUE)&(statsI$pass_mendel==TRUE)&(statsI$pass_imiss==TRUE)
statsI=cbind(statsI, ffI)

selectedI=statsI[which(statsI$ffI),1]
selectedI=cbind(statsI[which(statsI$ffI),2],selectedI)

ind_pass = sum(ffI, na.rm=T)
fail_imiss = sum(statsI$pass_imiss!=TRUE)
fail_het = sum(statsI$pass_het!=TRUE)
fail_mendel = sum(statsI$pass_mendel!=TRUE)
```

## Final statistics of pre-QC filtering
Starting number of individuals: `r totalI`  
Number of individuals passed all filters: `r ind_pass`  
Individuals flagged by IMISS filter: `r fail_imiss`  
Individuals flagged by HET filter: `r fail_het`  
Individuals flagged by MERR filter: `r fail_mendel`  

## Writing final output
Final output will now be written to directory `r outDir`. Files *full-stats* contain full filtering statistics, files *selected* contain the lists for automatic selection of individuals and SNPs to keep.

```{r final_out, echo=FALSE}
write.table(statsI, file=paste(outDir,"full-statsI.txt",sep=""), quote=F, sep="\t",row.names=F, col.names=T)
write.table(selectedI, file=paste(outDir,"selectedI.txt",sep=""), quote=F, sep="\t",row.names=F, col.names=F)
```